flowCore a Bioconductor package for high throughput flow cytometry

### 2008-12-12 paper/bmc_article.tex
some changes of the paper and addition of work flows
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@35870 bc3139a8-67e5-0310-9ffc-ced21a209358

% Load packages
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

@@

%Review style settings
\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
%\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
%% The computational tools we have developed are distributed in the R
%% software language as the Bioconductor package
%% \Rpackage{flowCore}. 
The Bioconductor package \Rpackage{flowCore} is a freely available,
highly functional, and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent
analysis. \Rpackage{flowCore} implements computationally efficient
data structures and a range of specialized methods addressing all
components of a typical FCM analysis work flow, including
compensation, transformation, and gating. \Rpackage{flowCore} runs on
Windows, Mac OS X, and Linux/Unix operating systems. Conceptually, the
underlying data structures that build the core of our software can be
implemented in any programming language. We chose the R language for
its advantages in statistical programming and visualization. 
%% In this paper, we describe the functionality and methods underlying
%% \Rpackage{flowCore}, and provide examples of their use.

@@

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML Candidate Recommendation
(CR), an emerging ISAC standard developed in collaboration with the
ISAC Data Standards Task Force, which makes it possibile to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

\subsection*{flowCore basic data structures}
\subsubsection*{flowFrame}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other Bioconductor packages. All
information from a single FCS file, i.e., the collection of events and
the accompanying metadata, is stored in one single container. We call
the structure that hold this data a \Rclass{flowFrame}. Raw data
values as well as associated metadata of a \Rclass{flowFrame} can be
accessed programmatically. Most commonly, the metadata consist of
descriptors of the stains used in the experiment and the respective
measurement channels, information about compensation performed at the
instrument side and any additional keywords the user deems to be
important to annotate the data. During the creation of a
\Rclass{flowFrame}, a number of quality checks are performed to ensure
data integrity.

@@

Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sup-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

@@

The ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, for example the proportions
of events falling in a single gate or in a combination of
gates. Complex combinations and hierarchies of gates can be captured
in objects of class \Rclass{filterSet}, allowing to apply multi-step
gating strategies. The definition of gates in \Rpackage{flowCore}
follows the Gating-ML CR, thus any \Rpackage{flowCore} gating strategy
can be reproduced by any other software that also adheres to the
standard.

@@

\begin{verbatim}
> library(flowCore)
> library(flowQ)
> qaReport(dat, c("qaProcess.timeline", "qaProcess.timeflow", 
                  "qaProcess.cellnumber"))
\end{verbatim}

@@

Kernel density estimation is a well-known problem in statistical
computing, and a lot of effort has been invested in the development of
good software to address it. The modular design of \Rpackage{flowCore}
allows to easily integrate these existing solutions into our
framework. In this example, we directly use R code from the
\Rpackage{feature} package developed by \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in \Rpackage{flowCore} is available to other package
authors, as we have seen with the afore-mentioned \Rpackage{flowClust}
package.

We can chose one of the many visualization option from the flowViz
package to plot the results of the recent filtering operation. A very
basic matrix of density plots is shown in Figure \ref{xyplot}, where
each panel in the matrix represents the data of one individual
patient.

@@

\subsection*{Workflow proposal(??)}

Florian?

@@

\section*{Figures}
  \subsection*{Figure 1 - flowCore Framework}
  \begin{figure}[hp]
    \centering
    \includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
    \caption{\label{fig1:FrameWork}{For each experiment, the content of
        the FCS files, phenotypic and metadata are stored in a
        \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
        corresponds to one FCS file. All basic operations (e.g.,
        compensation, transformation, gating) can be applied to either
        single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
  \end{figure}

  \subsection*{Figure 2 - Quality assessment}
    \begin{figure}[hp]
      \centering
      \includegraphics[width=0.98\textwidth]{flowQ.jpg}
      \caption{\label{flowQ}%
        HTML quality assessment report generated by the flowQ package for a
        subset of data from an experiment focussing on Graft-Versus-Host
        Disease \citep{brinkman2007hcf}. Rows correspond to the samples in
        the set, columns to different quality checks.}
    \end{figure}

    \subsection*{Figure 3 - Quality assessment}
    \begin{figure}[hp]
      \centering
      \includegraphics[width=0.75\textwidth]{xyplot}
      \caption{\label{xyplot}%
        Scatterplot matrix of a single \Rclass{flowSet} from an experiement
        focussing on immune tolerance following kidney
        transplantation. Outlines of the gating regions identified by an
        automated gating operation are added on top of the density
        representation of the data.}
    \end{figure}

@@

  \subsection*{Table 2 - Filter and gate classes}
  \begin{table}[ht]
    \begin{center}
      \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|c|}{Gates} \\
        \hline
        rectangleGate & n-dimensional rectangular regions \\
        quadGate & quadrant regions in two dimensions \\
        polygonGate & polygonal regions in two dimensions \\
        polytopeGate & generalization of polygon in n dimensions \\
        ellipsoidGate & n-dimensional ellipsoid region \\
        \hline
        \multicolumn{2}{|c|}{Filters} \\
        \hline
        sampleFilter & random sub-sampling\\
        expressionFilter & results of a boolean expression \\
        kmeansFilter & K-means clustering \\
        norm2Filter & bivariate normal distribution \\
        curv1Filter & local density regions in 1D \\
        curv2Filter & density regions in 2D \\
        timeFilter & abnormal data acquisition over time \\
        \hline
        filterSet & gating strategies \\
        \hline
      \end{tabular}
    \end{center}
  \end{table}

>>>>

% Load packages
%\usepackage[authoryear,round]{natbib}
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{graphicx}
%\usepackage{setspace}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

@@

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

@@

%%%%%%%%%%%%%%%%%%
\section*{Implementation}
%% The computational tools we have developed are distributed in the R
%% software language as the Bioconductor package
%% \Rpackage{flowCore}. 
The Bioconductor package \Rpackage{flowCore} is a freely available,
highly functional, and extensible FCM data analysis platform that
enables researchers to efficiently handle FC-HCS data and encourages
open development of tools for their coherent analysis. In our
implementation of \Rpackage{flowCore} we rely on two important lessons
learned from the field of gene expression data analysis: the first
being the importance of data structures that reflect the underlying
data and facilitate the manipulations that are of most interest, while
the second is the importance of a modular architecture that allows for
many developers to extend and use the underlying infrastructure and to
combine tools in complex work flows.  \Rpackage{flowCore} implements
such computationally efficient data structures and a range of
specialized methods addressing all components of a typical FCM
analysis work flow, including compensation, transformation, and
gating. \Rpackage{flowCore} runs on Windows, Mac OS X, and Linux/Unix
operating systems. Conceptually, the underlying data structures that
build the core of our software can be implemented in any programming
language. We chose the R language for its advantages in statistical
programming and visualization.
%% In this paper, we describe the functionality and methods underlying
%% \Rpackage{flowCore}, and provide examples of their use.

@@

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~\ref{fig1:FrameWork}).  Moreover, it is
a software implementation of the Gating-ML Candidate Recommendation
(CR), an emerging ISAC standard developed in collaboration with the
ISAC Data Standards Task Force, which makes it possible to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

\subsection*{flowCore basic data structures}
\subsubsection*{flowFrame}
\Rpackage{flowCore}'s primary task is the representation and basic
manipulation of FCM data. This is accomplished through a data model
very close to that adopted by other successful Bioconductor
packages. All information from a single FCS file, i.e., the collection
of events and the accompanying metadata, is stored in one single
container. We call the structure that hold this data a
\Rclass{flowFrame}. Raw data values as well as associated metadata of
a \Rclass{flowFrame} can be accessed programmatically. Most commonly,
the metadata consist of descriptors of the stains used in the
experiment and the respective measurement channels, information about
compensation performed at the instrument side and any additional
keywords the user deems to be important to annotate the data. During
the creation of a \Rclass{flowFrame}, a number of quality checks are
performed to ensure data integrity.

@@

Typically, the basic operations in FCM analyses adhere to the
following common work flow: the data need to be compensated (if that
was not already done on the instrument) and transformed, and
sub-populations of interest need to be selected based on a set of
(predominantly sequential) gates. All software solutions for FCM
analysis offer support for these operations, typically in an
interactive, GUI-driven interface. In \Rpackage{flowCore} we have
taken the approach to abstractly describe these operations and build a
set of tools to perform them on both \Rclass{flowFrames} and
\Rclass{flowSets}. While transformation, and to a certain extent
compensation, are fairly routine operations with only limited
potential for improvement, being able to implement new methodologies
for gating of FCM data, and extending the capabilities of
\Rpackage{flowCore} through object oriented programming are features
that clearly sets our framework apart from other FCM analysis
tools. By factoring out as much of the bookkeeping as possible,
programmers can focus on the actual operations rather then having to
deal with the tedious details of data integration and
access. Third-party methods can act on their own as first-class
citizens in the analysis framework, without breaking the work flow or
the basic infrastructure. This design allows for the straightforward
extension of \Rpackage{flowCore}'s capabilities, and has already
fostered the development of a number of valuable add-ons
\cite{lo2008agf,sarkar2008ufv}.

@@

The ability to programmatically access gates is a prerequisite for
automated or semi-automated gating. By utilizing a unified interface
for all different types of gates, the user is able to subset data sets
as well as to create summary statistics, for example the proportions
of events falling in a single gate or in a combination of
gates. Complex combinations and hierarchies of gates can be captured
in multi-step gating strategies. The definition of gates in
\Rpackage{flowCore} follows the Gating-ML CR, thus any
\Rpackage{flowCore} gating strategy can be reproduced by any other
software that also adheres to the standard and vice versa.

@@

\begin{verbatim}
> library(flowQ)
> qaReport(dat, c("qaProcess.timeline", 
                  "qaProcess.timeflow", 
                  "qaProcess.cellnumber"))
\end{verbatim}

@@

Kernel density estimation is a well-known problem in statistical
computing, and a lot of effort has been invested in the development of
good software to address it. The modular design of \Rpackage{flowCore}
allows to easily integrate these existing solutions into our
framework. In this example, we directly use R code from the
\Rpackage{feature} package \cite{wand2008}. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in \Rpackage{flowCore} is available to other package
authors.

We can chose one of the many visualization option from the
\Rpackage{flowViz} package to plot the results of the recent filtering
operation. A very basic matrix of density plots is shown in Figure
\ref{xyplot}, where each panel in the matrix represents the data of
one individual patient.

@@

Scripting languages like R provide a natural representation of work
flows through a sequence of code instructions in regular text
files. This allows for the rapid development and testing of new ideas,
however it is not very well suited for routine data analysis
tasks. Furthermore, the overhead of data management and variable
tracking can be considerable. To that end, \Rpackage{flowCore} also
provides data structures that help organize sequences of typical FCM
data analysis operations and complex gating strategies into concise
work flows. These structures are self-reflective, they contain all
intermediate results and offer a unified user interface to assess the
progress and the outcome of an anlysis.

@@

\section*{Figures}
  \subsection*{Figure 1 - flowCore Framework}
  \begin{figure}[hp]
    \centering
    \includegraphics{Figure1-flowCoreFrameWork.jpg}
    \caption{\label{fig1:FrameWork}{For each experiment, the content of
        the FCS files, phenotypic and metadata are stored in a
        \Rclass{flowSet}. Each \Rclass{flowFrame} in a \Rclass{flowSet}
        corresponds to one FCS file. All basic operations (e.g.,
        compensation, transformation, gating) can be applied to either
        single \Rclass{flowFrames} or a \Rclass{flowSet} simultaneously.}}
  \end{figure}

  \subsection*{Figure 2 - Quality assessment}
    \begin{figure}[hp]
      \centering
      \includegraphics{flowQ.jpg}
      \caption{\label{flowQ}%
        HTML quality assessment report generated by the flowQ package for a
        subset of data from an experiment focussing on Graft-Versus-Host
        Disease \cite{brinkman2007hcf}. Rows correspond to the samples in
        the set, columns to different quality checks.}
    \end{figure}

    \subsection*{Figure 3 - Quality assessment}
    \begin{figure}[hp]
      \centering
      \includegraphics{xyplot.pdf}
      \caption{\label{xyplot}%
        Scatterplot matrix of a single \Rclass{flowSet} from an experiement
        focussing on immune tolerance following kidney
        transplantation. Outlines of the gating regions identified by an
        automated gating operation are added on top of the density
        representation of the data.}
    \end{figure}

@@

  \subsection*{Table 2 - Filter and gate classes}
  \begin{table}[ht]
    \label{table2}
    \begin{center}
      \begin{tabular}{|l|l|}
        \hline
        \multicolumn{2}{|c|}{Gates} \\
        \hline
        rectangleGate & n-dimensional rectangular regions \\
        quadGate & quadrant regions in two dimensions \\
        polygonGate & polygonal regions in two dimensions \\
        polytopeGate & generalization of polygon in n dimensions \\
        ellipsoidGate & n-dimensional ellipsoid region \\
        \hline
        \multicolumn{2}{|c|}{Filters} \\
        \hline
        sampleFilter & random sub-sampling\\
        expressionFilter & results of a boolean expression \\
        kmeansFilter & K-means clustering \\
        norm2Filter & bivariate normal distribution \\
        curv1Filter & local density regions in 1D \\
        curv2Filter & density regions in 2D \\
        timeFilter & abnormal data acquisition over time \\
        \hline
        filterSet & gating strategies \\
        \hline
      \end{tabular}
    \end{center}
  \end{table}
